---
title: "Lab 3: MDS and CA practical"
format: pdf
editor: visual
author: Leonel Nabaza and Michael Jiménez
---

## Third deliverable

**Instructions**

Work to be done in pairs. Deadline: 1st of April.

Like in the previous deliverables, please give precise answers with proper justifications.

**Introduction**

In this last exercise we are going to perform a little of the two last methods explained at class: Multidimensional Scaling and Correspondence Analysis.

Maybe, during these exercises we learn a secret relationship between these methods...

## MDS and CA

### MDS

Given the following matrix of **squared** **distances**:

```{r}
D <- read.csv("https://raw.githubusercontent.com/Afrojam/toy_datasets/refs/heads/main/economy_race_distances.csv", row.names = 1)
n <- nrow(D)
ones <- rep(1, n)
print(D)
```

**Compute**:

1.  The double centering matrix $B$ .
2.  Perform eigendecomposition on $B$ . Then deliver:
    1.  The eigenvalues. Comment if there are strong negative eigenvalues.

    2.  Justify if the provided distances come from a metric.

    3.  Compute the goodness of fit and chose the number of principal coordinates. We want to have a quality of the approximation greater than 80%.

    4.  Calculate the new coordinates in the principal coordinates space and perform a scatter plot.

    5.  Provide a plausible label for the principal coordinates based on the position of the points

```{r}
# double centering matrix
D <- as.matrix(D)
storage.mode(D) <- "double"

n <- nrow(D)
J <- diag(n) - matrix(1, n, n) / n
B <- -0.5 * J %*% D %*% J

round(B, 4)
```

```{r}
#eigen decomposition on B
eig <- eigen(B)
eig$values
```

There aren't strong negative eigenvalues, since they are raised to very small powers.

Since the negative eigenvalues are negligible and the matrix $B$ is positive semidefinite up to numerical precision, the provided squared distances are compatible with a Euclidean metric.

For the goodness of the fit we shall use:

$$
\frac{\sum_{i = 1}^k \lambda_i}{\sum_{\lambda_j > 0} \lambda_j}
$$

```{r}
lambda <- eig$values
lambda[lambda < 0]
lambda_pos <- lambda[lambda > 1e-10]
gof <- cumsum(lambda_pos) / sum(lambda_pos)

data.frame(
  Dimension = 1:length(lambda_pos),
  Eigenvalue = round(lambda_pos, 6),
  Cumulative_GoF = round(gof, 6)
)

k <- which(gof > 0.80)[1]
k
```

Although with one coordinate we'd have enough, it would be convenient for visual purposes to utilize two components.

For the new coordinates $X = V_k \Lambda_k^{1/2}$

```{r}
V <- eig$vectors[, 1:k, drop = FALSE]
X <- sweep(V, 2, sqrt(lambda[1:k]), `*`)

rownames(X) <- rownames(D)
round(X, 4)
```

For the plot, we need to use two dimensions:

```{r}
k_plot <- 2
X2 <- eig$vectors[, 1:k_plot, drop = FALSE] %*% diag(sqrt(lambda[1:k_plot]), nrow = k_plot)
rownames(X2) <- rownames(D)

X2_df <- data.frame(
  Dim1 = X2[,1],
  Dim2 = X2[,2],
  Point = rownames(X2)
)

library("ggplot2")
library("ggrepel")

ggplot(X2_df, aes(x = Dim1, y = Dim2, label = Point)) +
  geom_point(size = 3) +
  geom_text_repel() +
  theme_minimal() +
  labs(title = "MDS map using the first two principal coordinates",
       x = "Principal coordinate 1",
       y = "Principal coordinate 2")
```

A plausible interpretation of the principal coordinates is the following:\
\
- **Principal coordinate 1** seems to represent the main racial contrast in the data, separating **African American** (and to a lesser extent **Hispanic**) from **White** and **Asian**.\
- **Principal coordinate 2** seems to capture a more specific contrast that mainly separates the **Hispanic** group from the other categories.\
\
Therefore, a plausible labeling is:\
\
- **Dim 1**: African American / Hispanic vs White / Asian contrast\
- **Dim 2**: Hispanic differentiation

### CA

Now, we are going to use the following data from the library **openintro.**

```{r}
library(openintro)
data(hsb2)
tab <- table(hsb2$ses, hsb2$race)
```

The data, as we are going to use it, consists in a cross table between two categorical variables:

-   **Ses**: Socio economic status of student's family. With categories:

    -   low, middle and high.

-   **Race:** Student's race, with categories:

    -   african american, asian, hispanic and white.

**Exercises:**

1.  Give the correspondence table (P), the row profiles matrix, the column profiles matrix and the average row profile (r) and the average column profile (c).

    ```{r}
    N <- as.matrix(tab)
    n_total <- sum(N)

    P <- N / n_total
    r <- rowSums(P)
    c <- colSums(P)

    Dr_inv <- diag(1 / r)
    Dc_inv <- diag(1 / c)

    row_profiles <- Dr_inv %*% P
    col_profiles <- P %*% Dc_inv

    cat("Correspondence table P:\n")
    print(round(P, 6))

    cat("\nRow masses r:\n")
    print(round(r, 6))

    cat("\nColumn masses c:\n")
    print(round(c, 6))

    cat("\nRow profiles matrix:\n")
    print(round(row_profiles, 6))

    cat("\nColumn profiles matrix:\n")
    print(round(col_profiles, 6))
    ```

2.  Give a ternary plot, if it is possible. If not, justify why. If yes, explain what you observe about the relationship between points and vertex.

    A ternary plot is only possible if the composition has exactly 3 parts.\
    Here:

    1.   Row profiles have 4 components (because race has 4 categories), so they **cannot** be represented in a ternary plot.

    2.   Column profiles have 3 components (because ses has 3 categories), so they **can** be represented in a ternary plot.

    ```{r}
    cat("Number of row profile components:", ncol(row_profiles), "\n")
    cat("Number of column profile components:", nrow(col_profiles), "\n")

    install.packages("ggtern")
    library(ggtern)

      # Column profiles: each race gives a distribution over ses = low, middle, high
      col_prof_df <- as.data.frame(t(col_profiles))
      col_prof_df$Race <- rownames(col_prof_df)

      # Rename columns to something simpler if needed
      colnames(col_prof_df)[1:3] <- c("low", "middle", "high")

      ggtern(data = col_prof_df, aes(x = low, y = middle, z = high, label = Race)) +
        geom_point(size = 3) +
        geom_text_repel() +
        labs(title = "Ternary plot of column profiles (Race over SES)") +
        theme_bw()
    ```

3.  Give the matrix of standardized residuals (S).

    ```{r}
    Dr_mhalf <- diag(1 / sqrt(r))
    Dc_mhalf <- diag(1 / sqrt(c))

    S <- Dr_mhalf %*% (P - r %*% t(c)) %*% Dc_mhalf

    cat("Standardized residuals matrix S:\n")
    print(round(S, 6))
    ```

4.  Obtain the CA solution. Then deliver:

    -   Principal inertias.

    -   Quality of representation with two principal axis.

    -   Correlations and contributions.

    ```{r}
    svd_S <- svd(S)

    U <- svd_S$u
    Delta <- diag(svd_S$d)
    V <- svd_S$v

    singular_values <- svd_S$d
    principal_inertias <- singular_values^2

    cat("Singular values:\n")
    print(round(singular_values, 6))

    cat("\nPrincipal inertias:\n")
    print(round(principal_inertias, 6))
    ```

    ```{r}
    qual_2d <- sum(principal_inertias[1:2]) / sum(principal_inertias)
    cat("Quality of representation with two axes:", round(qual_2d, 6), "\n")

    Phi <- Dr_mhalf %*% U
    Gamma <- Dc_mhalf %*% V

    F <- Phi %*% Delta
    G <- Gamma %*% Delta

    rownames(F) <- rownames(N)
    rownames(G) <- colnames(N)

    cat("Row principal coordinates:\n")
    print(round(F, 6))

    cat("\nColumn principal coordinates:\n")
    print(round(G, 6))
    ```

    ```{r}
    #similarity for rows
    row_ctr <- matrix(0, nrow = nrow(F), ncol = length(principal_inertias))
    col_ctr <- matrix(0, nrow = nrow(G), ncol = length(principal_inertias))

    for (a in seq_along(principal_inertias)) {
      row_ctr[, a] <- r * F[, a]^2 / principal_inertias[a]
      col_ctr[, a] <- c * G[, a]^2 / principal_inertias[a]
    }

    rownames(row_ctr) <- rownames(N)
    rownames(col_ctr) <- colnames(N)
    colnames(row_ctr) <- paste0("Dim", seq_along(principal_inertias))
    colnames(col_ctr) <- paste0("Dim", seq_along(principal_inertias))

    cat("Row contributions:\n")
    print(round(row_ctr, 6))

    cat("\nColumn contributions:\n")
    print(round(col_ctr, 6))


    #similarity for columns
    row_cos2 <- F^2 / rowSums(F^2)
    col_cos2 <- G^2 / rowSums(G^2)

    rownames(row_cos2) <- rownames(N)
    rownames(col_cos2) <- colnames(N)
    colnames(row_cos2) <- paste0("Dim", seq_len(ncol(row_cos2)))
    colnames(col_cos2) <- paste0("Dim", seq_len(ncol(col_cos2)))

    cat("Row cos²:\n")
    print(round(row_cos2, 6))

    cat("\nColumn cos²:\n")
    print(round(col_cos2, 6))
    ```

5.  Perform a row profiles biplot. What you observe?

    ```{r}
    rownames(F) <- rownames(N)
    rownames(Gamma) <- colnames(N)

    row_plot_df <- data.frame(
      Dim1 = F[, 1],
      Dim2 = F[, 2],
      Label = rownames(F),
      Type = "Row"
    )

    col_std_df <- data.frame(
      Dim1 = Gamma[, 1],
      Dim2 = Gamma[, 2],
      Label = rownames(Gamma),
      Type = "Column (standard)"
    )

    library(ggplot2)
    library(ggrepel)

    ggplot() +
      geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
      geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
      geom_point(data = row_plot_df, aes(Dim1, Dim2), size = 3) +
      geom_text_repel(data = row_plot_df, aes(Dim1, Dim2, label = Label)) +
      geom_point(data = col_std_df, aes(Dim1, Dim2), shape = 17, size = 3) +
      geom_text_repel(data = col_std_df, aes(Dim1, Dim2, label = Label)) +
      labs(
        title = "CA Row Profiles Biplot",
        x = "Dimension 1",
        y = "Dimension 2"
      ) +
      theme_minimal()
    ```

    From the CA row profiles biplot, several patterns can be observed:\
    \
    - Dimension 1 represents a strong contrast between racial groups. African American (and to a lesser extent Hispanic) are located on the positive side, while White is on the negative side. This dimension reflects a gradient opposing African American vs White populations.\
    \
    - Dimension 2 mainly separates the Hispanic group from the others, as it is located far above all other categories. This suggests a specific pattern for Hispanic that is not captured by the main axis.\
    \
    - Regarding the association between race and socioeconomic status:\
     - African American is strongly associated with low SES.\
     - White is associated with high SES.\
     - Asian is closer to middle SES, indicating a more balanced profile.\
     - Hispanic appears more distinct and not strongly aligned with a single SES category.

6.  Perform a column profiles biplot. What you observe?

    ```{r}
    rownames(Phi) <- rownames(N)
    rownames(G) <- colnames(N)
    row_std_df <- data.frame(
      Dim1 = Phi[, 1],
      Dim2 = Phi[, 2],
      Label = rownames(Phi),
      Type = "Row (standard)"
    )

    col_plot_df <- data.frame(
      Dim1 = G[, 1],
      Dim2 = G[, 2],
      Label = rownames(G),
      Type = "Column"
    )

    ggplot() +
      geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
      geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
      geom_point(data = col_plot_df, aes(Dim1, Dim2), size = 3) +
      geom_text_repel(data = col_plot_df, aes(Dim1, Dim2, label = Label)) +
      geom_point(data = row_std_df, aes(Dim1, Dim2), shape = 17, size = 3) +
      geom_text_repel(data = row_std_df, aes(Dim1, Dim2, label = Label)) +
      labs(
        title = "CA Column Profiles Biplot",
        x = "Dimension 1",
        y = "Dimension 2"
      ) +
      theme_minimal()
    ```

    From the CA column profiles biplot, we observe the following:\
    \
    - Dimension 1represents a clear socioeconomic gradient, separating high SES (left) from low SES (right), with middle SES located closer to the center.\
    \
    - There is a strong association between:\
     - Low SES and African American\
     - High SES and White\
     - Middle SES and Asian\
    \
    - The Hispanic group is located near the origin, indicating a more balanced distribution across SES categories.\
    \
    - Dimension 2 mainly distinguishes the middle SES category from the others, as it is located much higher than both low and high SES.\
    \
    Overall, the plot confirms a strong relationship between race and socioeconomic status, with a dominant gradient from high to low SES along the first dimension.

7.  Interpret the symmetric biplot. Use visual information to give contributions and the relative frequency for each point.

    ```{r}
    Delta_half <- diag(sqrt(singular_values), nrow = length(singular_values))

    F_sym <- Phi %*% Delta_half
    G_sym <- Gamma %*% Delta_half

    rownames(F_sym) <- rownames(N)
    rownames(G_sym) <- colnames(N)

    row_sym_df <- data.frame(
      Dim1 = F_sym[, 1],
      Dim2 = F_sym[, 2],
      Label = rownames(F_sym),
      Set = "Rows"
    )

    col_sym_df <- data.frame(
      Dim1 = G_sym[, 1],
      Dim2 = G_sym[, 2],
      Label = rownames(G_sym),
      Set = "Columns"
    )

    library(ggplot2)
    library(ggrepel)

    ggplot() +
      geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
      geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
      geom_point(data = row_sym_df, aes(Dim1, Dim2), size = 3) +
      geom_text_repel(data = row_sym_df, aes(Dim1, Dim2, label = Label)) +
      geom_point(data = col_sym_df, aes(Dim1, Dim2), shape = 17, size = 3) +
      geom_text_repel(data = col_sym_df, aes(Dim1, Dim2, label = Label)) +
      labs(
        title = "CA Symmetric Biplot",
        x = "Dimension 1",
        y = "Dimension 2"
      ) +
      theme_minimal()
    ```

    From the symmetric biplot, we observe the overall association structure between race and socioeconomic status.

    The first dimension represents the main gradient in the data. On the positive side, **African American** is closely associated with **low SES**, while on the negative side, **White** is associated with **high SES**. This indicates a strong socioeconomic contrast linked to racial categories.

    The second dimension captures a secondary pattern. The **Hispanic** group and the **middle SES** category are both located higher along this axis, suggesting a specific association that is not explained by the main gradient.

    The **Asian** group is located near the origin, indicating a more balanced profile and weaker association with extreme SES categories.

    Overall, the symmetric biplot confirms that the dominant relationship in the data is the opposition between **African American / low SES** and **White / high SES**, with a secondary structure involving **Hispanic** and **middle SES**.

8.  Can you made the connection between MDS and CA? Explain the cause and which results should be equal.

    Correspondence Analysis (CA) can be interpreted as a particular case of Multidimensional Scaling (MDS). Specifically, CA is equivalent to applying classical MDS to the row (or column) profiles using the chi-square distance.

    The reason for this connection is that both methods rely on a spectral decomposition of a matrix that represents the relationships between observations. In MDS, this is done through the eigendecomposition of the double-centered distance matrix. In CA, it is done through the singular value decomposition of the standardized residual matrix.

    The key equivalences are:

    -   The eigenvalues in MDS correspond to the principal inertias in CA (i.e., the squared singular values).
    -   The coordinates obtained in both methods are the same up to a possible change of sign or rotation.
    -   The geometric structure (distances between points) is preserved in both representations.

    Therefore, CA can be seen as MDS applied to profiles under the chi-square metric, adapted specifically for contingency tables.

**Note**: There are no restrictions on the libraries to use, but some steps maybe only possible with manual calculation (i.e.: not using libraries that automatize the steps).
