---
title: "CA_step_by_step"
author: "Jamie Arjona"
date: "2025-03-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Canonical Correlation Analysis
Analysis of the two way table obtainable by comparing two categorical variables.
Input: Given a two-way table (contingency table)
Goal: Visualization and interpretation of relationships between rows and columns. 
How categories are related based on their deviations from independence.
Can be interpreted as PCA for categorical variables.

## Data
We have 5 different species of insects:
Ladybugs, caterpillars, weevils, aphids and spiders
From 3 different public parks in the city:
A, B and C.
Populations of bugs are represented in the following two way table:

```{r}
ladybugs <- c(38,28,29)
caterpillars <- c(113,123,85)
weevils <- c(116,91,90)
aphids <- c(80,46,57)
spiders <- c(38,24,42)
data <- matrix(c(ladybugs, caterpillars, weevils, aphids, spiders), nrow=3)
tdata <- as.table(data)
park_names <- c("A","B","C")
bug_names <- c("ladybugs", "caterpillars", "weevils", "aphids", "spiders")
rownames(tdata) <- park_names
colnames(tdata) <- bug_names
tdata
N <- tdata # notation
I <- nrow(tdata)
J <- ncol(tdata)
```

**Goal**:
Are there differences in the distributions of the insects across the public parks?

*Note*: We can represent our data as the number of dimension for CA will be I-1=2
from min(I-1, J-1).

## Obtain correspondence matrix and profiles matrices
Correspondence matrix: Matrix of probabilities.
Profile: Vector that sums up to 1 (also know as frequency vectors)
```{r}
ones_r <- rep(1,I)
ones_c <- rep(1,J)
n <- t(ones_r) %*% N %*% ones_c
n
n <- as.integer(n)
```

Correspondence matrix:
```{r}
P <- N / n
P
# sanity check
t(ones_r) %*% P %*% ones_c
r <- P%*%ones_c # row masses
c <- t(P)%*% ones_r # column masses
r
c
# diagonal matrices for later
Dr <- diag(c(r))
Dc <- diag(c(c))
```

Profile matrices
```{r}
# Row profiles and its column marginal
R <- solve(Dr)%*%P
rownames(R) <- park_names
# Column profiles and its row marginal
C <- solve(Dc)%*%t(P)
C <- t(C)
colnames(C) <- bug_names

R
R %*% ones_c # row profiles so its row marginals are 1
C
ones_r %*% C # column profiles so its column marginals are 1
# sanity check (masses (marginals from P) as weighted averages of the profiles)
all.equal(t(r)%*%R, t(c))
all.equal(t(ones_r)%*%P, t(c))

all.equal(t(c)%*%t(C), t(r))
all.equal(t(ones_c)%*%t(P), t(r))
```
We can calculate the contingency ratio. It indicates how many times the entry is
greater or smaller than the respective masses. It must be the same for row profile and
column profile.
```{r}
all.equal(C[1,1] / r[1], R[1,1] / c[1])
solve(Dr)%*%C
R%*%solve(Dc)
```

## Ternary plots
Visualization of the ternary plot. Ternary plots work when the cumulative sum of
the elements in a vector is 1, e.g.: compositional data or profile vectors.
They make easy to interpret:
- Distances between points: Relations between categories of both cat. variables.
- Inertia: How spread are the points wrt to the center. Low inertia means independence
in the chi-square test for independence.
It works for C and R as have marginals that sum up to 1 (rows for C and columns 
for R).
Reshape the data so we have as columns the cateogrical variable with 3 categories.

```{r}
# install.packages("ggtern")
library(ggtern)
df_C <- data.frame(A=C[1,], B=C[2,], C=C[3,], lab=bug_names)
df_R <- data.frame(A=R[1,], B=R[2,], C=R[3,], lab=bug_names)

ggtern(data = df_C, aes(x = A, y = B, z = C)) +
  geom_point(shape=1, size=4, color="red") +
  geom_text(aes(label = 1:nrow(df_C))) +
  theme_rgbw() + 
  theme_showarrows() +
  theme_clockwise() +
  ggtitle("Columns profile") + 
  xlab("A") +
  ylab("B") +
  zlab("C")

df_R <- data.frame(A=R[1,], B=R[2,], C=R[3,], lab=bug_names)
ggtern(data = df_R, aes(x = A, y = B, z = C)) +
  geom_point(shape=1, size=4, color="red") +
  geom_text(aes(label = 1:nrow(df_R))) +
  theme_rgbw() + 
  theme_showarrows() +
  theme_clockwise() +
  ggtitle("Columns profile") + 
  xlab("A") +
  ylab("B") +
  zlab("C")
```
In the ternary plots we can show the massess and the coordiantes of the points.
We can think of the average profile as a special point as it is the centroid of 
a weighted average (rows or columns) and it is situated at the center of the 
profiles but more attracted to those with more weight. 
Observe that the average of the data will be always at the center of the ternary 
plot, i.e.: c(1/3,1/3,1/3).
In general, centroid or barycenter is an alternative term for the weighted average.
In Correspondence Analysis the weighted average gives the average profile and the
weights used to compute the weighted average is what is call the mass. E.g.:
In the row profile, we have the row margin as the masses and the column margin
as the average row profile (it is the result of dividing each row by its corresponding
row marginal and sum it up).

Next we show that in column profiles, the row marginal represent the average 
column profile:
```{r}
C_ext <- cbind(C,r)
colnames(C_ext) <- c(bug_names, "r_avg")
df_C_ext <- data.frame(A=C_ext[1,], B=C_ext[2,], C=C_ext[3,], lab=c(bug_names, "r_avg"))
gp <- ggtern(data = df_C_ext, aes(x = A, y = B, z = C, color = lab)) +
  geom_point(shape = ifelse(df_C_ext$lab == "r_avg", 4, 1), size = 3) + 
  geom_text(aes(label = lab), vjust = -0.5) + 
  scale_color_manual(values = c("r_avg" = "red")) +
  theme_rgbw() +
  theme_showarrows() +
  theme_clockwise() +
  ggtitle("Columns profile") +
  xlab("A") +
  ylab("B") +
  zlab("C") +
  theme(legend.position = "one")
gp
```


## Chi-square distances
Used to compute variation on the data of a contingency table.
The key concept in Correspondence Analysis is the use of the Chi-square distance.
The Chi-square distance is a weighted Euclidean distance. The weights are the 
inverse of the masses so a low mass component indicates a rare category and has
a higher contribution.
In the Chi square test of homogeneity, independence (homogeneity) means that the 
discrepancies between observed and expected frequencies are small. If this does 
not holds, there is a dependence and this is measure through a test statistic
(the chi square statistic) that follows a chi square distribution.
chi square distance = sum (observed - expected)^2 / expected
The geometric interpretation is considering the concept of inertia: how far are
observations from the average.
inertia = sum_i (ith row mass) (chi square distance from ith profile to average profile)^2
Inertia will be high when row profiles have large deviatinos from their average 
profile and low if close.
Inertia can be also considered as a correlatin between categories of the variables:
The higher the inertia, the closer the row profiles to the vertices of a ternary
plot, i.e.: closer the categories between the categorical variables.
Interpretation of minimum and maximum inertia:
Minimum inertia is 0: All profiles are equal to their respective average profile.
Maximum inertia is equal to the number of dimensions of the space profile (remember 
that is min(I-1, J-1)).

First, the chi-square statistic as a measure of deviations from the average using
the contingency table.
Chi-square test review:
The chi square test measures the deviations from the means (row or column masses) 
of the cells. 
Each cell contributes towards the chi square statistic. 
The chi square statistic follows a chi square distribution with(I-1)x(J-1) degrees of freedom.
```{r}
# In R, just one call:
chisq.test(N)

# Chi square statistic computation for rows (we need to recover original counts):
Er <- matrix((c*n) %o%r, ncol = J, byrow = TRUE)
Ec <- matrix((r*n) %o%c, ncol = J)
all.equal(Er,Ec) # (c[j]/n)*r[i] == c[j]*(r[i]/n)
chisq_contributions <- ((N-Er)**2/Er)
round(chisq_contributions,3)
chisq = sum(chisq_contributions)
dof <- (I-1)*(J-1)
pval <- 1-pchisq(chisq, df=8)

# Turns out that the chi square statistic uses a weighted euclidean distance
# called the chi square distance and that what we do is calculate the inertia,
# the deviations (distance of each cell) from the corresponding mean.

# define a function for computing the chi square distance
squared_chi_square_distance <- function(x, y, m) {
  if (length(x) != length(y)) {
    stop("Vectors must have the same length.")
  }
  # Avoid division by zero by adding a small constant
  epsilon <- 1e-10

  chi_sq <- 0
  for (i in 1:length(x)) {
    chi_sq <- chi_sq + ((x[i] - y[i])**2 / (m[i] + epsilon))
  }
  chi_sq <- as.numeric(chi_sq)
  return(chi_sq)
}

total_inertia_by_row <- function(Q, cq, rq) {
  # expects a row profiles matrix and the average row and column profile.
  # TODO: generalize to input N and add argument byrows.
  
  if (sum(rowSums(Q))!=nrow(Q)) {
    stop("Not a row profiles matrix.")
  }
  
  tot_inertia <- 0
  for (i in 1:nrow(Q)){
    tot_inertia <- tot_inertia + rq[i] * squared_chi_square_distance(Q[i,], cq, cq)
  }
  return(tot_inertia)
}

total_inertia_by_col <- function(Q, cq, rq) {
  # expects a row profiles matrix and the average row and column profile.
  # TODO: generalize to input N and add argument byrows.
  
  if (sum(colSums(Q))!=ncol(Q)) {
    stop("Not a row profiles matrix.")
  }
  
  tot_inertia <- 0
  for (i in 1:ncol(Q)){
    tot_inertia <- tot_inertia + cq[i] * squared_chi_square_distance(Q[,i], rq, rq)
  }
  return(tot_inertia)
}

# sanity check
# library(vegan)
# all.equal(vegdist(N, method="chisq")[1], sqrt(squared_chi_square_distance(R[1,], R[2,], c)))


tot_inerR <- total_inertia_by_row(R, c, r)
tot_inerC <- total_inertia_by_col(C, c, r)
all.equal(tot_inerR, chisq / n)
all.equal(tot_inerC, chisq / n)
```

Now, chi square distance with profiles
```{r}
# Distances between column profiles
sqrt(squared_chi_square_distance(C[,1],C[,1], r))
sqrt(squared_chi_square_distance(C[,1],C[,2], r))
sqrt(squared_chi_square_distance(C[,1],C[,3], r))
sqrt(squared_chi_square_distance(C[,1],C[,4], r))
sqrt(squared_chi_square_distance(C[,1],C[,5], r))

# Distances between row profiles
sqrt(squared_chi_square_distance(R[1,],R[1,], c))
sqrt(squared_chi_square_distance(R[1,],R[2,], c))
sqrt(squared_chi_square_distance(R[1,],R[3,], c))
```



### Example visualization of the inertia
Visualize that the inertia concept is analogous to spread of the categories.
```{r}
table1 <- matrix(c(21,20,19,20,21,19,20,19,20,19,19,22,20,21,20), nrow = 5, byrow = TRUE)
colnames(table1) <- c("A", "B", "C")
rownames(table1) <- 1:5
table2 <- matrix(c(23,18,19,16,24,20,24,17,18,17,19,18,20,22,15), nrow = 5, byrow = TRUE)
colnames(table2) <- c("A", "B", "C")
rownames(table2) <- 1:5
table3 <- matrix(c(27,15,13,13,30,14,29,15,12,16,18,45,15,22,16), nrow = 5, byrow = TRUE)
colnames(table3) <- c("A", "B", "C")
rownames(table3) <- 1:5
table4 <- matrix(c(30,11,10,10,34,11,34,12,10,12,10,57,14,33,12), nrow = 5, byrow = TRUE)
colnames(table4) <- c("A", "B", "C")
rownames(table4) <- 1:5

table1
table2
table3
table4
```
As ternary plots:
```{r}
library(gridExtra)

plot_tables <- function(data_table, title){
  df <- as.data.frame(data_table)
  df$id <- rownames(df)
  gp <- ggtern(data = df, aes(x = A, y = B, z = C)) +
  geom_point() +
  geom_text(aes(label = id), vjust = 1) +
  ggtitle(title)+
  xlab("A") +
  ylab("B") +
  zlab("C")
  gp
}


gp1 <- plot_tables(table1, "Inertia ")
gp2 <- plot_tables(table2, "Inertia ")
gp3 <- plot_tables(table3, "Inertia ")
gp4 <- plot_tables(table4, "Inertia ")
grid.arrange(gp1,gp2, gp3, gp4)
```
In the upper left plot the inertia is small, i.e.: variables are independent.
In the bottom right plot the inertia is high, i.e.: variables are homogeneous.
In the extreme case of maximal inertia we can consider that the variables are the
same.

## The correspondence analysis
Now we are ready to perform correspondence analysis.
The **ca** package performs simple correspondence analysis given a contingency 
table N or correspondence table P.
It will perform all calculations and computations explained in the slides and we
can access all of them easily.
Also it allows easy visualization of biplots.
```{r}
library(ca) # simple correspondence analysis
res <- ca(P)
print(res)
summary(res)
res$sv # Why we have only two?
res$nd
res$rowinertia
res$colinertia
res$rowcoord
res$colcoord
res$rowmass
res$colmass
res$rowdist # are squared chi square distances, they need to be transformed by sqrt
res$coldist # are squared chi square distances, they need to be transformed by sqrt
```

Biplots:
```{r}
plot.ca(res, map="rowprincipal", arrows=c(FALSE,TRUE), main="Row profiles biplot", mass=TRUE)
plot.ca(res, map="colprincipal", arrows=c(TRUE,FALSE), main="Column profiles biplot", mass=TRUE)
plot.ca(res, map="symmetric", arrows=c(FALSE,FALSE), main="Symmetric biplot", mass=TRUE)
```

We can also check the contributions by color gradient:
```{r}
plot.ca(res, map="symmetric", arrows=c(FALSE,FALSE), mass=TRUE, contrib = "relative")
```

We can add supplementary individuals:
```{r}
wasps <- c(0.7,0.2,0.1)
Dpark <- c(0.1,0.2,0.04,0.05,0.25,0.3)
P.ext <- cbind(P, wasps)
P.ext <- rbind(P.ext, Dpark)
res.ext <- ca(P.ext, supcol=6, suprow=4)
plot.ca(res.ext, map="symmetric", arrows=c(FALSE,FALSE))
```

