---
title: "Factor Analysis"
author: "Jamie Arjona"
date: "2026-03-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(psych)
library(GGally)
```



# The FA model
Code that shows how to compute the factor model using R different libraries.

## Data
Data consists on exams grades in differents subjects. In some of the exams, students
were able to use the course book (marked as O).
```{r}
data <- read.table("data/FA_BookOpenClosed.dat")
cnames <- c("mechanics (C)", "vectors (C)", "algebra (O)", "analysis (O)", "statistics (O)")
colnames(data) <- cnames
head(data)
summary(data)
ggpairs(data, progress=FALSE)
X <- as.matrix(data)
```

## Factor analysis in base R: factanal function
The *factanal* function from STATS package (included in R) performs maximum likelihood
estimate of the FA model. We can choose to do different rotations, (e.g. varimax,
promax, etc.). The scores are calculated by Bartlett method or regression 
(Thurstone/Thomson).
```{r}
# ?factanal
res.ml <- factanal(X, factors=2, rotation="none", scores="Bartlett")
print(res.ml)

```

Understanding loadings:
```{r}
res.ml$loadings
# communalities: the link between variable variance and factor contribution to it
diag(res.ml$loadings %*% t(res.ml$loadings)) 
# If entries of loadings matrix get squared then the row marginal are the loadings
# and the column marginals are the factors variance.
colSums(res.ml$loadings**2) # communalities
1-res.ml$uniquenesses
rowSums(res.ml$loadings**2) # sum of squares of loadings



# We want to recover the correlation matrix with few variables
(LLt <- res.ml$loadings%*%t(res.ml$loadings))
cor(X)
LLt +diag(res.ml$uniquenesses)
```

The obtained loadings are ugly and not interpretable at all, we can do better
just rotating the loadings
```{r}
res.ml$loadings
varimax(res.ml$loadings)
res.ml.varimax <- factanal(X, factors=2, rotation="varimax", scores="Bartlett")
print(res.ml.varimax$loadings)

# communalities, sum of squared loadings and uniqueness do not change
res.ml.varimax$loadings %*% t(res.ml.varimax$loadings)
LLt
all.equal(res.ml.varimax$uniquenesses, res.ml$uniquenesses)

```

## fa method from psych library
This library implements other methods to estimate FA beyond maximum likelihood and
also permits to use other rotations such as **oblimin**.

**Note**: the complexity, *com* column, should not be confused with the communalities.
```{r}
#?fa
res.fa <- fa(X, nfactors=2, rotate="none", scores="Bartlett")
print(res.fa)
res.fa.varimax <- fa(X, nfactors=2, rotate="varimax", scores="Bartlett")
print(res.fa.varimax)
res.fa.oblimin <- fa(X, nfactors=2, rotate="oblimin", scores="Bartlett")
print(res.fa.oblimin)
```



Observe that **oblimin* returns three matrices:
```{r}
(P <- unclass(res.fa.oblimin$loadings))
(S <- unclass(res.fa.oblimin$Structure))
(Phi <- unclass(res.fa.oblimin$Phi))
(T_mat <- res.fa.oblimin$rot.mat)

print("Structure matrix equal to matrix multiplication of simple structure with correlations:")
all.equal(P%*%Phi, S)

print("Correlation matrix computed from rotation matrix:")
manual_phi <- solve(t(T_mat)%*%T_mat)
Phi
manual_phi
```

We want the p-value to be large, so we don't reject the null hypothesis
```{r}
res.fa$PVAL
res.fa.varimax$PVAL
res.fa.oblimin$PVAL
```


Thurstone factor scores:
```{r}
res.fa.thurstone <- fa(X, nfactors=2, rotate="none", scores="Thurstone")
W <- scale(X)%*%solve(cor(X))

all.equal(head(W%*%res.fa.thurstone$loadings), head(res.fa.thurstone$scores))


```

Bartlett factor scores
```{r}
psi <- diag(res.fa.varimax$uniquenesses)
W.pre <- solve(psi)%*%res.fa.varimax$loadings
W.post <- solve(t(res.fa.varimax$loadings)%*%solve(psi)%*%res.fa.varimax$loadings)
W <- W.pre %*% W.post

all.equal(head(scale(X)%*%W),head(res.fa.varimax$scores))
```

We can use the Scorings for ANOVA or regression tasks:
```{r}
head(res.fa.oblimin$scores)
res.fa.oblimin$score.cor
# how well factor scores represent theoretical latent factors, want those to be
# as close as possible to 1:
res.fa.oblimin$r.scores
```

We can explore the scores in a biplot:
```{r}
biplot(res.fa, 
       main = "Biplot of Factor Analysis NO rotation",
       col = c("blue", "red"),
       pch = 16,
       cex = 0.8)
biplot(res.fa.varimax, 
       main = "Biplot of Varimax Factor Analysis",
       col = c("blue", "red"),
       pch = 16,
       cex = 0.8)
biplot(res.fa.oblimin, 
       main = "Biplot of Oblimin Factor Analysis",
       col = c("blue", "red"),
       pch = 16,
       cex = 0.8)
```


## The PCA method
We can compute the loadings using the PCA (warning: not all FA conditions are
met)
```{r}
Xs <- scale(X, center=TRUE, scale=TRUE)
n <- nrow(Xs)
R <- 1/(n-1)* t(Xs)%*%Xs
res.pca <- eigen(R)
D <- sqrt(diag(res.pca$values))
m <- 2
L <- res.pca$vectors[,1:m]%*%D[1:2,1:m]
colnames(L)<- cnames[1:m]
SSL <- diag(t(L)%*%L)
gof <- sum(diag(D)[1:m])/sum(diag(D))
# j$loadings
comms <- rowSums(L**2)
specific <- 1 - comms # since variances are scaled they are 1
SSL
round(gof,3)
comms
specific

# Compare with previous methods
cor(Xs)
L%*%t(L)+diag(specific)
LLt + diag(res.ml$uniquenesses)
# P%*%Corrs%*%t(P) + Psi
res.fa.oblimin$loadings%*%res.fa.oblimin$Phi%*%t(res.fa.oblimin$loadings) + diag(res.fa.oblimin$uniquenesses)

# sanity check: Sum of squared loadings, loadings, specificity
# res.pca.check <- principal(R, nfactors=2, rotate="none")
# sum of squared loadings: 
# res.pca.check$values
# loadings:
# res.pca.check$loadings
# specificity
# res.pca.check$uniqueness
```

The varimax command from STATS (base R) performs the varimax rotation on the loadings
matrix:
```{r}
res.pca.rot <- varimax(L)
print(res.pca.rot$loadings)
print(res.pca.rot$rotmat)

# sanity check principal(R, nfactors=2, rotate="varimax")
```

They key is to understand that FA is a **causal model**:
```{r}
fa.diagram(res.fa.varimax)
fa.diagram(res.fa.oblimin)
```

