---
title: "fda examples"
author: "Jamie Arjona"
date: "2024-05-02"
output: pdf_document
---
Libraries
```{r}
library(psych) # all fa methods
library(readr)
library(tidyverse)
```


```{r}
data <- read_csv("data/expenses.csv")
row_names <- data %>% select(1) %>% as.list
X <- data %>% select(bread, vegetables, fruits,  meat, poultry,  milk,  wine) %>% as.matrix
n <- dim(X)[1]
p <- dim(X)[2]
```
```{r}
library(calibrate) # biplot
```


Let's perform FA without rotation.
```{r}
out.ml <- factanal(X, factors=2, rotation="none", scores="Bartlett") # MLE
print(out.ml)
```

Biplot:
```{r}
ze <- rep(0, p)
plot(out.ml$loadings,type="n",xlim=c(-1,1),ylim=c(-1,1))
text(out.ml$loadings,labels=colnames(X), col="blue")
arrows(ze, ze, out.ml$loadings[,1], out.ml$loadings[,2], length = 0.07,col="blue")
abline(h=0,v=0,col="gray")
circle(1)
```



Next, we are going to try to interpret the data from the rotated factors:
```{r}
out.rotated <- factanal(X, factors=2, rotation="varimax", scores="Bartlett")
print(out.rotated)
```


Biplot:
```{r}
ze <- rep(0, p)
plot(out.rotated$loadings,type="n",xlim=c(-1,1),ylim=c(-1,1))
text(out.rotated$loadings,labels=colnames(X), col="blue")
arrows(ze, ze, out.rotated$loadings[,1], out.rotated$loadings[,2], length = 0.07,col="blue")
abline(h=0,v=0,col="gray")
circle(1)
```

# Using Psych library
The psych library offers the *fa* function which provides five different methods 
to perform factor analysis: minimum residual, principal factor, weighted least squares,
generalized least squares and maximum likelihood factor analysis.

**principal** gives us FA from PCA
```{r}
fa.principal <- principal(X, nfactors= 2, rotate="varimax", scores=TRUE)
print(attributes(fa.principal))
```
And the results are:
```{r}
print(fa.principal$loadings)
print(fa.principal$scores)

print(fa.principal$communality) # sum of squares of loadings for each variable
print(fa.principal$uniquenesses) # 1 - communality for each variable.
print(fa.principal$rot.mat)

```
Using Principal Factor analysis
```{r}
fa.pfa <- fa(X, nfactors= 2, rotate="varimax", scores=TRUE, fm="pa")
print(attributes(fa.pfa))
```
Be careful, the library warns us about a Heywood case (statistic assumptions not 
satisfied, i.e.: negative variances or correlation estimates greater than 1)

```{r}
print(fa.pfa$loadings)
print(fa.pfa$scores)

print(fa.pfa$communality) # sum of squares of loadings for each variable
print(fa.pfa$uniquenesses) # 1 - communality for each variable.
print(fa.pfa$rot.mat)
```
Using psych with maximum likelihood estimates
```{r}
fa.psych.ml <- fa(X, nfactors= 2, rotate="varimax", scores=TRUE, fm="ml")
print(attributes(fa.psych.ml))
```

And the results:
```{r}
print(fa.psych.ml$loadings)
print(fa.psych.ml$scores)

print(fa.psych.ml$communality) # sum of squares of loadings for each variable
print(fa.psych.ml$uniquenesses) # 1 - communality for each variable.
print(fa.psych.ml$rot.mat)
```
# Interpretation of the data
In general, the two factors are making a distinction about milk and bread vs fruits, meat and poultry.
The first factor makes a distinction of wine vs the rest of the variables so we can
interpret this factor as occasional use vs daily use. The problem comes from the
fact that wine is not well represented with two factors (should we add a third factor?)
The second factor becomes evident once one uses the CHILDREN variable and becomes clear
that the expanses on bread and milk is closely related to the number of childrens.

# Exercise
Replicate the results from principal of the psych library starting with eigen on 
the covariance (or correlation) matrix and varimax from the default package stats.
Take into account that you will need to perform an inverse matrix calculation 
(solve or ginv functions) and multiply some matrices.