---
title: "fa_example_2"
author: "Jamie Arjona"
date: "2024-05-15"
output: pdf_document
---
# Exploring rotations from PCA
Practitioners usually perform the varimax rotation after obtaining the PCA solution
in order to get better and clear interpretability of the results.

This is not FA, but is a "mix" between PCA and FA.

```{r}
library(FactoMineR)
library(factoextra)
library(readr)
library(tidyverse)
```

```{r}
data <- read_csv("data/expenses.csv")
row_names <- data %>% select(1) %>% as.list
X <- data %>% select(bread, vegetables, fruits,  meat, poultry,  milk,  wine) %>% as.matrix
n <- dim(X)[1]
p <- dim(X)[2]
rownames(X) <- row_names$...1
```

Performing PCA:
```{r}
pca.res <- PCA(X, ncp = 4, graph = FALSE)
pca.ind <- pca.res$ind
pca.var <- pca.res$var
```

A fast summary of the results:
```{r}
summary(pca.res)
```

## Variables:
**Observation**: Notice that we are in the variable space. Here the distances are represented by
the Mahalanobis distance, so coordinates of the variables are correlations (because
standarization) between variables and principal axis. In addition, notice that 
the square of the correlations (or coordinates) are the squared cosines as distances
are measure in Mahalanobis distance.
```{r}
pca.res$var
```
We can get extra interpretability using the factoextra package. For example a plot
of the contributions and square cosines of the variables:
```{r}
fviz_contrib(pca.res, "var", axes=1)
fviz_cos2(pca.res, "var", axes=1)
```
We can also obtain information about the relation of which variables are more related
to the principal components performing some statistical tests:
```{r}
dimdesc(pca.res)
```


## Individuals
Output from the individuals.
NOTE: Distance is computed as:
sqrt(rowSums((scale(as.matrix(X)) * sqrt(nrow(X)/(nrow(X) - 1)))**2))
gives us the length of the vector of the standardized individual. Is not related
to the coordinate in the PC space but gives us some intuitions of proximity
between individuals in the original space.

Distances in the individual space are Euclidean.
```{r}
pca.res$ind
```

Plots of contributions and squared cosines:
```{r}
fviz_contrib(pca.res, "ind", axes=1)
fviz_cos2(pca.res, "ind", axes=1)
```
It seems that there is a size effect caused by the individual number 12.

## Biplot

Check the individuals biplot:
```{r}
fviz_pca_biplot(pca.res, repel=TRUE)
```

Our goal is to perform varimax in order to obtain better interpretability:
```{r}
nd <- 2
Y_var <- pca.res$var$coord[,1:nd]
Y_ind <- pca.res$ind$coord[,1:nd]
Y_var.rot <- varimax(Y_var)

rot.eigenval <- diag(t(Y_var.rot$loadings)%*%Y_var.rot$loadings)
```
Observe that the name of the items have changed to loadings once we performed the
rotation.
In this case (unlike in FA, PCA does not computes specificities) sum of diagonal 
elements of matrix of loadings is equal to the sum of the eigenvalues.
```{r}
sum(rot.eigenval)
sum(pca.res$eig[1:nd,1])
```

We can obtain the rotated individuals from the rotation estimated on the variables
```{r}
Xs <- scale(X) #needed
Y_stan.ind.rot <- Xs%*%solve(cor(X))%*%Y_var.rot$loadings
Y_ind.rot <- Y_stan.ind.rot %*% diag(sqrt(rot.eigenval)) # undo standardization
```

Performing the plots
```{r}
library(calibrate)
ze = rep(0,p)
plot(Y_var.rot$loadings,type="n",xlim=c(-1,1),ylim=c(-1,1))
text(Y_var.rot$loadings, col="blue", labels = colnames(X))
arrows(ze, ze, Y_var.rot$loadings[,1], Y_var.rot$loadings[,2], length = 0.07,col="blue")
abline(h=0,v=0,col="gray")
circle(1)
plot(Y_ind.rot,type="n")
text(Y_ind.rot,col=as.numeric(data$STATUS), labels = rownames(X))
abline(h=0,v=0,col="gray")
```
Compare with the not rotated version:
```{r}
ze = rep(0,p)
plot(Y_var,type="n",xlim=c(-1,1),ylim=c(-1,1))
text(Y_var, col="blue", labels = colnames(X))
arrows(ze, ze, Y_var[,1], Y_var[,2], length = 0.07,col="blue")
abline(h=0,v=0,col="gray")
circle(1)
plot(Y_ind,type="n")
text(Y_ind,col=as.numeric(data$STATUS), labels = rownames(X))
abline(h=0,v=0,col="gray")
```

