---
title: "mds_example_1"
author: "Jamie Arjona"
date: "2024-05-06"
output: pdf_document
---

Load libraries
```{r}
library("MASS") # isoMDS and shepard plot.
```


```{r}
Spain <- as.matrix(read.table("data/SpainDist.dat", header=TRUE))
rownames(Spain) <- colnames(Spain)
n <- nrow(Spain)
mds.out <- cmdscale(Spain,k=n-1,eig=TRUE)
X <- mds.out$points[,1:2]
plot(X[,2],X[,1],type="n", xlab="", ylab="", main="Map of Spain",asp=1,
xlim=c(-800,800),ylim=c(-800,500))
points(X[,2],X[,1],pch=19,cex=0.5)
text(X[,2],X[,1],rownames(Spain), cex=0.5,pos=1)
```

```{r}
ev <- mds.out$eig
gof <- mds.out$GOF
print(round(ev,digits=2))
print(round(gof,digits=4))
```


```{r}
(ev[1]+ev[2])/sum(abs(ev))

(ev[1]+ev[2])/sum(ev[ev>0])
```

```{r}
mds.out <- cmdscale(Spain,k=2,eig=TRUE)
mds.out$GOF
```

Sheppard plot, quality of the representation
```{r}
plot(as.dist(Spain), dist(mds.out$points), ylab = "approximated distances", 
     xlab="true distances", xlim=c(0,1500), ylim=c(0,1500), main="Shepard plot")
abline(lm(dist(mds.out$points) ~ as.dist(Spain)), col="red")
```

## Spain distances non-metric MDS approach:
```{r}
init_config <- scale(matrix(runif(n*2), ncol=2), scale = FALSE)
nmds.out <- isoMDS(as.dist(Spain), y=init_config, k=2, maxit=200)
Y <- nmds.out$points
```

Next we are going to use as initial config the solution obtained by PCorA:
```{r}
nmds2.out <- isoMDS(as.dist(Spain), y=X, k=2, maxit=200)
Y2 <- nmds.out$points
```

Nonmetric MDS solution:
Note: Sometimes axes are reversed and need to be adjusted using multiplication by -1.
```{r}
plot(1*Y[,2],-1*Y[,1],pch=19, asp=1, main="Map of Spain using non metric MDS")
text(1*Y[,2], -1*Y[,1], rownames(Spain), cex=0.5,pos=1)
```
Shepard plot of NMDS:
```{r}
plot(as.dist(Spain), dist(Y), ylab = "approximated distances", 
     xlab="true distances", main="Shepard plot")
abline(lm(dist(Y) ~ as.dist(Spain)), col="red")
```

In the ndms with initial as cmdscale output:
```{r}
plot(as.dist(Spain), dist(Y2), ylab = "approximated distances", 
     xlab="true distances", main="Shepard plot")
abline(lm(dist(Y2) ~ as.dist(Spain)), col="red")
```

Plotting the stress for different number of dimensions:
```{r}
stress_list <- c()
for (k in 1:5){
  init_config <- scale(matrix(runif(n*k), ncol=k), scale = FALSE)
  stress_list <- append(stress_list, nmds.out <- isoMDS(as.dist(Spain), y=init_config, k=k, maxit=200)$stress)
}
plot(stress_list, main="Stress versus dimensionality", xlab="# of dimensions", 
     ylab="Stress value", pch=19)
lines(stress_list)
```
## Without distances (rank orders)
NMDS does not need distances, it just needs the ordered ranks of the data, i.e.:
data ordered by rank similarity. With that it can obtain good approximates of the data.
Now we are going to create a new dataset called SpainRank in which we will have the
rank orders between cities based on the distance instead of the true distances.
```{r}
#?rank
# rank(c(10, 11, 100, 35, 1))
init <- (matrix(runif(n*2),ncol=2))
Spain.out <- isoMDS(Spain,y=init,k=2,maxit=100)
X.original <- Spain.out$points # backup the obtained points

DistVec <- Spain[lower.tri(Spain)]
DistVecRank <- rank(DistVec)
SpainRank <- Spain
SpainRank[lower.tri(SpainRank)] <- DistVecRank
SpainRank[upper.tri(SpainRank)] <- 0
diag(SpainRank) <- 0
SpainRank <- SpainRank + t(SpainRank)
SpainRank.out <- isoMDS(SpainRank,y=init,k=2,maxit=100)
X.rank <- SpainRank.out$points
```
Let's compare the outputs:
```{r}
pairs(cbind(X.original,X.rank))
cor(cbind(X.original,X.rank))
```

Comparing MDS and NMDS solutions
```{r}
pairs(cbind(X.original,mds.out$points[,1:2]), labels = c("nmds x", "nmds y", "mds x", "mds y"))
cor(cbind(X.original,mds.out$points[,1:2]))
```


# Toy example: Metric vs non-metric
```{r}
toy <- matrix(c(0,1,10,1,0,9,10,9,0), nrow=3)
toy.rank <- matrix((rank(toy)), nrow = 3)
diag(toy.rank) <- 0
mds.out <- cmdscale(dist(toy), k=1)
nmds.out <- isoMDS(toy.rank, y=matrix(runif(3*1), nrow=3), k=1)
plot(mds.out, c(1,1,1), xlab = "First dim", ylab="Second dim", asp=1)
plot(nmds.out$points, c(1,1,1), xlab = "First dim", ylab="Second dim", asp=1)
```

## Soda perception example
We can work with subjective similarities or perceptions as distances and use metric
or nonmetric mds. This gives us a way to interpret this type of data when we relate
it with the axes.
```{r}
soda <- matrix(c(
  0, 35, 79, 86, 76, 63, 58, 62, 66, 26,
  35, 0, 54, 56, 30, 41, 86, 81, 24, 61,
  79, 54, 0, 70, 51, 38, 78, 72, 69, 70,
  86, 56, 70, 0, 66, 90, 50, 89, 66, 89,
  76, 30, 51, 66, 0, 35, 76, 68, 23, 63,
  63, 41, 38, 90, 35, 0, 77, 54, 35, 68,
  58, 86, 78, 50, 76, 77, 0, 66, 77, 59,
  62, 81, 72, 89, 68, 54, 66, 0, 71, 34,
  66, 24, 69, 66, 23, 35, 77, 71, 0, 59,
  26, 61, 70, 89, 63, 68, 59, 34, 59, 0
), nrow = 10)
soda_names <- c("Diet Pepsi", "RC", "Yukon", "Dr.Pepper", "Shasta", "Coca Cola", "Diet Dr.Pepper", "TAB", "Pepsi", "Diet Rite")
rownames(soda) <- soda_names
colnames(soda) <- soda_names
# sanity check isSymmetric.matrix(data)
```

```{r}
mds.soda <- cmdscale(soda, k=2, eig=T)
mds.soda$eig
```

Plot of the new configuration:
```{r}
plot(X[,2],X[,1],type="n", xlab="", ylab="", main="Soda perception",asp=1,
xlim=c(-40,40),ylim=c(-40,40), axes = "F")
# axis(1, pos=0)
# axis(2, pos=0)
arrows(c(0,c(-40,40)[1]),c(c(-40,40)[1],0),c(0,c(-40,40)[2]),c(c(-40,40)[2],0),0.05)
points(mds.soda$points[,2],mds.soda$points[,1],pch=19,cex=0.5,)
text(mds.soda$points[,2],mds.soda$points[,1],soda_names, cex=0.8,pos=1)
points(0,0, pch=19)
text(0,0, "(0,0)", cex=0.8,pos=1)
```

# A look to another package: vegan
This package has gained some popularity recently as it provides a NMDS with more
automation than the one given by *isoMDS* of the MASS package.
Vegan offers the function *metaMDS* that calls monoMDS (monotonic MDS) or isoMDS
but also automatizes the initialization of the init configuration of points, performs
scaling and rotation. In addition, we can use different distances like bray-curtis (check vegdist help)
and use auxiliary functions to interpret the output in a similar fashion as one does with PCA.

Note: Because *vegan* is made for ecological data, the nomenclature is somehow special.
It makes the distinction of `sites` and `species`. For us, the sites are the samples
and species are the variables. We do not need species as we are working with distances/similarities.
```{r}
library("vegan")
# ?metaMDS
# ?vegdist
```

We are going to use the soda data, as it is based on perception and we can obtain
insights from the axis:
```{r}
mds.vegan.soda <- metaMDS(soda, k=2, eig=T)
```

Let's check the output:
1. googness gives us goodness of fit for each point.
2. stressplot gives us the Shepard diagram of the nonmetric mds. Since it uses
monotonic regression, it is no longer a stright line, but as in metric mds, we
are interested in obtain a diagonal cloud of points.
```{r}
goodness(mds.vegan.soda)
stressplot(mds.vegan.soda)
```

Plotting the output
```{r}
plot(mds.vegan.soda, "sites", type="n")
orditorp(mds.vegan.soda, "sites")
arrows(c(0,c(-40,40)[1]),c(c(-40,40)[1],0),c(0,c(-40,40)[2]),c(c(-40,40)[2],0),0.05)
```
