---
title: "pca_factominer-2026"
author: "Jamie Arjona"
date: "2026-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(FactoMineR)
library(factoextra)
library(GGally)
```

## USArrests PCA on Covariance
```{r USArrests}
data("USArrests")
head(USArrests)
X <-USArrests
ggpairs(X, progress = FALSE) +
  theme_minimal()

```

Covariance matrix:
```{r}
n <- nrow(X)
p <- ncol(X)
Xc <- scale(x=X, center=TRUE, scale=FALSE)
(S <- 1/(n-1) * t(Xc)%*%Xc)

```

PCA
```{r}
?PCA
res1.pca <- PCA(Xc, scale.unit=FALSE, ncp=p, graph=FALSE)
```

### Eigenvectors and eigenvalues interpretations
```{r}
res1.pca$eig
```




## PCA on correlation matrix
```{r}
Xs <- scale(Xc, center=TRUE, scale=TRUE)
res2.pca <- PCA(Xs, scale.unit=TRUE, ncp=p, graph=FALSE)
```

### Eigenvectors and eigenvalues interpretations

#### Eigenvalues
```{r}
res2.pca$eig
diag(res2.pca$svd$vs)
```
Kaiser Rule
```{r}
threshold <- mean(res2.pca$eig[,1])
eig.idx <- res2.pca$eig[,1] > threshold
mean(res2.pca$eig[eig.idx,1])
```

Scree plot + Kaiser rule + elbow
```{r}
fviz_eig(res2.pca, addlabels = TRUE) + 
  geom_hline(yintercept = (1/ncol(Xc)*100), linetype = "dashed", color = "red")
```
#### Eigenvectors
```{r}
res2.pca$svd$V
```


#### Individuals analysis
```{r}
Z <- res2.pca$ind$coord
head(Z)
head(Xs%*%res2.pca$svd$V[,1])
res2.pca$ind$coord/Xs%*%res2.pca$svd$V
sqrt((n)/(n-1))
```

Relationships original variable and Scores
```{r}
cov(X, Z)
cor(X, Z)
```

Marginals:
```{r}
col_marginals <- colSums(Z**2)
row_marginals <- rowSums(Z**2)
1/n*col_marginals
row_marginals
```


Contributions:
```{r}
head(round(res2.pca$ind$contrib,4))
head(round(Z**2 %*% diag(1/col_marginals)*100,4))
```

Potential outliers:
```{r}
threshold <- colMeans(res2.pca$ind$contrib)
# Individuals with large contribution in any dimension
out.any.idx <- which(apply(res2.pca$ind$contrib, 1, function(x) any(x>threshold)))
round(res2.pca$ind$contrib[out.any.idx,],4)

# Individuals with large contribution in specific dimension
out.specific.idx <- which(res2.pca$ind$contrib[,1]>threshold)
round(res2.pca$ind$contrib[out.specific.idx,],4)

# Individuals with large contribution in all dimensions
out.all.idx <- which(apply(res2.pca$ind$contrib, 1, function(x) all(x>threshold)))
round(res2.pca$ind$contrib[out.all.idx,],4)

# Individuals with small contribution in all dimensions
noout.all.idx <- which(apply(res2.pca$ind$contrib, 1, function(x) all(x<threshold)))
round(res2.pca$ind$contrib[noout.all.idx,],4)


```
Visualizations of contributions
```{r}
fviz_contrib(res2.pca, "ind", axes = 1)
fviz_contrib(res2.pca, "ind", axes = 2)
```

Quality of representation
```{r}
head(round(res2.pca$ind$cos2,4))
head(round(diag(1/row_marginals)%*%Z**2,4))
```

Visualization of cos2
```{r}
fviz_cos2(res2.pca, "ind", axes = 1)
fviz_cos2(res2.pca, "ind", axes = 2)
```

Approximation with k components
```{r}
k <- 2
apply(res2.pca$ind$cos2[,1:k], 1, sum)*100

# In general
apply(res2.pca$ind$cos2, 1, cumsum)*100
```

#### Variables analysis
```{r}
L <- res2.pca$svd$V%*% diag(res2.pca$svd$vs)
res2.pca$var$cor
res2.pca$var$coord
cor(X,Z)
L

```

Marginals
```{r}
row_marginals <- rowSums(L**2)
col_marginals <- colSums(L**2)
```


Contributions of variable to each component
```{r}
apply(L, 2, function(x) x**2/sum(x**2))*100
res2.pca$var$contrib
```

Visualization of contributions of the original variables to the PC
```{r}
fviz_contrib(res2.pca, "var", axes = 1)
fviz_contrib(res2.pca, "var", axes = 2)
```


Quality of representation
```{r}
res2.pca$var$cos2
res2.pca$var$cor**2
diag(1/row_marginals) %*%L**2
```

Representation with k components
```{r}
apply(res2.pca$var$cos2, 1, cumsum)*100
```
Visualization of quality of representations
```{r}
fviz_cos2(res2.pca, "var", axes = 1)
fviz_cos2(res2.pca, "var", axes = 2)
```


#### Visualizations

Individuals (Z scatterplot)
```{r}
fviz_pca_ind(res2.pca)
fviz_pca_ind(res2.pca, 
             col.ind = "cos2",
             alpha.ind = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
fviz_pca_ind(res2.pca, 
             axes = c(1,3),
             col.ind = "cos2",
             alpha.ind = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```

Variables (L)

```{r}
fviz_pca_var(res2.pca)
fviz_pca_var(res2.pca,
             col.var = "cos2",
             alpha.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
fviz_pca_var(res2.pca,
             axes = c(1,3),
             col.var = "cos2",
             alpha.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
```

Biplot
```{r}
fviz_pca_biplot(res2.pca,
)
fviz_pca_biplot(res2.pca,
                col.ind = "cos2",
                gradient.cols = "RdYlBu",
                
                col.var = "contrib",
                alpha.var = "contrib",
                
                repel=TRUE
)

fviz_pca_biplot(res2.pca,
                axes=c(1,3),
                col.ind = "cos2",
                gradient.cols = "RdYlBu",
                
                col.var = "contrib",
                alpha.var = "contrib",
                
                repel=TRUE
)
             
```

