---
title: "Lab 2: PCA"
format: pdf
editor: visual
authors: Leonel Nabaza and Michael Jiménez
execute: 
  warning: false
  message: false
---

# Exercise 2 MVA

Work to be done in pairs.

Due date 22/03/2026.

## PCA

```{r}
#| include: false
library(GGally)
library(FactoMineR)
library(factoextra)
```

For this laboratory we are going to work with data about Italian restaurants in Manhattan.

The following instruction will load the data into the data variable\
(url to download the data:\
<https://raw.githubusercontent.com/Afrojam/toy_datasets/refs/heads/main/nyc.csv>\
)

```{r}
data = read.csv("https://raw.githubusercontent.com/Afrojam/toy_datasets/refs/heads/main/nyc.csv", row.names = 1)
head(data)
```

The variables are:

-   **Price**: Average price of the meal plus a drink, in US \$.

-   **Food, Decor, Service**: Food quality, local decoration and service as ranked by Zagat.com (0 to 25)

-   **Location**: Categorical variable: East if restaurant located on the East side and West if located on the West side.

Take a first look onto the data (consider the following plot when doing the analysis):

```{r}
ggpairs(data)
```

From now on, we will only work with numerical data, so we drop the Location variable.

```{r}
data2 <- data[,1:4]
```

### Exercises

For this exercise, libraries that automate the computation of PCA are permitted: prcomp from base R, PCA from FactoMineR, rda in Vegan, etc. It is also possible to use auxiliary libraries to perform the plots (i.e: factoextra)

1.  Perform a PCA on the data without scaling the variables. Show the principal components and their variance.

    ```{r}
    # pca on the data without scaling the variables. show principal components and their variance
    res.pca.noscale <- PCA(data2, scale.unit = FALSE, graph = FALSE)
    get_eigenvalue(res.pca.noscale)
    fviz_eig(res.pca.noscale, addlabels = TRUE, choice = "variance", axes = 1 )
    ```

    ```{r}
    fviz_eig(res.pca.noscale, addlabels = TRUE, choice = "eigenvalue", axes = 1 )
    ```

2.  Study the variance. Is there a size effect? If there is any, take action to correct it and from now on use the corrected principal components to perform the analysis. Justify your decisions.

    There a size in effect actually. If we inspect the plot we can see that PC1 explains 92.3% of the variance, while the other components explain 4.1, 2.9 and 0.7% respectively. This is evidence that there is a variable with a much greater variance and that it's dominating de PCA. Inspecting the dataset, we can suspect that the causing variable is "Price", since it usually achieves bigger values compared to the other variables. To correct this effect, the variables must be standardized before applying PCA. Therefore, from this point on, the analysis should be based on the PCA computed with scaled variables.

    ```{r}
    # Comparing the variance of variables
    diag(cov(data2))

    res.pca <- PCA(data2, scale.unit = TRUE, graph = FALSE)
    res.pca$eig
    ```

3.  Perform a scree plot and the accumulated variance captured by the principal components. Will be a good approximation if we only use the first two principal components?

    ```{r}
    res.pca <- PCA(data2, scale.unit = TRUE, graph = FALSE)
    #scree plot
    fviz_eig(res.pca, addlabels = TRUE)
    ```

    ```{r}
    get_eigenvalue(res.pca)
    ```

    The scree plot shows that the first principal component explains 74.92% of the variance, and the second explains 14.04%. Together, the first two components capture 88.96% of the total variance.\

    Since this percentage is very high, using only the first two principal components provides a good approximation of the data while significantly reducing the dimensionality. The scree plot also shows a clear elbow after the second component, indicating that the remaining components contribute relatively little additional information.

4.  Study the loadings (principal components with its standard deviation). What can be observed?

    ```{r}
    res.pca$var$coord
    ```

    The loadings show how each variable contributes to the principal components.\
    For the first principal component, all variables (Price, Food, Decor and Service) have large positive loadings. This indicates that PC1 represents the overall quality level of the restaurants: restaurants with higher values in this component tend to have higher prices and better ratings in food, decor and service.

    The second principal component contrasts Decor and Price with Food and Service. Decor and Price have positive loadings, while Food and Service have negative loadings. This suggests that PC2 separates restaurants that emphasize decor and price from those that emphasize food quality and service.

    The third and fourth components explain a small proportion of the variance, so they are less important for interpreting the structure of the data.

5.  Calculate the contributions and the squared cosines of the variables. Which variable contributed the most to the first principal component and which one to the second principal component? Which variables are better represented by the third and fourth principal components? Justify all your answers.

    ```{r}
    res.pca$var$contrib
    ```

    The variable that contributes the most to the first principal component is **Service** (27.11%). This indicates that the first principal component is strongly influenced by the service rating, although all variables contribute significantly.

    The variable that contributes the most to the second principal component is **Decor** (40.54%), followed by **Food** (37.59%). This shows that the second component mainly contrasts restaurants according to their decor and food ratings.

    ```{r}
    res.pca$var$cos2
    ```

    ```{r}
    fviz_contrib(res.pca, choice = "var", axes = 1)
    ```

    ```{r}
    fviz_contrib(res.pca, choice = "var", axes = 2)
    ```

    The variable best represented by the third principal component is **Price**, since it has the highest squared cosine value (0.153). This means that PC3 captures part of the variability associated with the price variable.

    For the fourth principal component, the variable **Service** has the highest squared cosine (0.061), meaning it is the variable best represented by this component.

6.  Calculate the PCA scores matrix, and the contributions and squared cosines of the individuals. What we can say from the restaurant Rao's and Gennaro in the context of the third principal component and its relation with the original data? Provide a restaurant that will have a poor representation if using only the first two principal components. Justify all your answers.

    ```{r}
    # PCA scores (coordinates of individuals)
    scores <- res.pca$ind$coord

    # Contributions of individuals
    contributions <- res.pca$ind$contrib

    # Squared cosines of individuals
    cosines <-res.pca$ind$cos2
    ```

    ```{r}
    scores["Rao's",]
    contributions["Rao's",]
    cosines["Rao's",]
    scores["Gennaro",]
    contributions["Gennaro",]
    cosines["Gennaro",]


    ```

    The third principal component is mainly associated with the **Price** variable. Rao's has a relatively large score on the third component (−1.47) and a high squared cosine (0.515), indicating that this restaurant is strongly characterized by the third principal component and therefore by the price variable.

    Gennaro also has a relatively large score on PC3 (−1.66), but its squared cosine is higher for the second component (0.467), meaning that its position is more strongly explained by PC2 than by PC3.

    Rao's would have a poor representation if only the first two principal components were used. The sum of its squared cosines for PC1 and PC2 is only about 0.48, which indicates that less than half of its variability is captured in the two-dimensional PCA representation. Therefore, the third component plays an important role in describing this restaurant.

7.  Perform a plot of the variables. What can we say about the principal axes and the variables?

    ```{r}
    fviz_pca_var(res.pca,
                 col.var = "black",
                 repel = TRUE)
    ```

    The variable plot (correlation circle) shows that all variables have strong positive loadings on the first principal component, which explains 74.9% of the variance. This indicates that the variables Price, Food, Decor and Service are positively correlated. Therefore, the first principal axis represents the overall quality level of the restaurants, where higher prices tend to be associated with better ratings in food, decor and service.

    The second principal axis (14% of the variance) contrasts Decor with Food and Service. Decor has a positive loading on this axis, while Food and Service have negative loadings. This suggests that the second component differentiates restaurants that emphasize decoration from those that emphasize food quality and service.

    The variables Food and Service are very close in the plot, indicating a strong positive correlation between them. All variables are well represented in the PC1–PC2 plane, as their vectors are long and close to the correlation circle.

8.  Perform a plot of the observations. What we can say about the following restaurants:

    1.  Veronica.

    2.  Ranbow Grill.

    3.  Lamarca.

    Remember to use all the information available about the representation in the reduced dimensional space.

    ```{r}
    fviz_pca_ind(
      res.pca,
      select.ind = list(name = c("Veronica", "Rainbow Grill", "Lamarca")),
      geom = c("point","text"),
      repel = TRUE
    )
    ```

    The first principal component represents the overall quality of the restaurants, while the second component contrasts decor with food and service. Veronica and Lamarca appear on the negative side of the first principal component, indicating lower values in the main variables such as price, food, decor and service compared with the average. Rainbow Grill is located on the positive side of the first component, suggesting a higher overall quality. In addition, its positive value on the second component indicates a stronger emphasis on decor compared with food and service.\

    ```{r}
    #| eval: false
    #| include: false
    res <- PCA(data2)
    restaurant <- c("Paola's","Sette Mezzo","Torre di Pisa","Primavera", "Veronica", "Rainbow Grill", "Fino", "Lamarca", "Basta Pasta")
    subset <- res$ind$coord[restaurant,]
    plot(subset, xlim = c(-5,5), ylim=c(-3,3), xlab = "PC1", ylab = "PC2")
    text(subset, labels = rownames(subset), pos = 4, cex = 0.7)
    abline(v=0, h=0)
    plot(scale(data[restaurant,c(4,3)], scale = FALSE), xlab = "SERVICE", ylab = "DECOR")
    text(scale(data[restaurant,c(4,3)], scale = FALSE), labels = rownames(subset), pos = 4, cex = 0.7)
    abline(v=0, h=0)
    ```

9.  We used location as supplementary variable and performed the following plot of individuals marked by their Location (points East, Triangles West):\
    \
    Based on your interpretation of the principal components, what can we say about East and West restaurants?

    ```{r echo=FALSE}
    res2 <- PCA(data, quali.sup="Location", graph=FALSE)
    fviz_pca_ind(res2, select.ind = list(cos2=35),habillage = "Location", palette=c("#00AFBB", "#E7B800"), xlim=c(-4,4), ylim=c(-2,2), mean.point=FALSE, repel=TRUE)
    ```

    Restaurants located in the **West** tend to appear on the positive side of the first principal component, which corresponds to higher overall quality, price, and ratings. Many of them also have positive values on the second component, suggesting a stronger emphasis on decoration.

    In contrast, **East** restaurants tend to be located on the negative side of the first component, indicating lower overall scores and prices compared with West restaurants. They are also more frequently associated with negative values of the second component, suggesting a relatively stronger emphasis on food and service rather than decoration.

10. One of your best friends is going to Manhattan and she wants to eat at a restaurant with the following conditions:\
    High food quality.\
    Not caring about the decoration.\
    Good service.\
    Cheap.\
    Which restaurant would you recommend? Justify your answer.

    Based on the PCA plot, the restaurant that best satisfies the desired conditions is **Tony's Di Napoli**. The first principal component represents the overall level of the restaurant, including price. Since the restaurant is located on the negative side of Dim1, it indicates a relatively lower price. The second principal component contrasts decoration with food and service; Tony's Di Napoli lies on the negative side of Dim2, suggesting stronger emphasis on food quality and service rather than decoration. Therefore, this restaurant matches the requirements of good food, good service, low concern for decor, and relatively low price.

    ## Factor Analysis

    The dataset *attitude* represents a survey of clerical employees (clerks) in 30 large financial institutions, measuring their "attitude" toward their workplace. The data are aggregated from the questionnaires of the approximately 35 employees for each of 30 randomly selected departments. The numbers give the percent proportion of favorable responses to seven questions in each department. The variables are:

    -   Rating: (Target variable). Overall rating of the department by its employees. It's how happy or satisfied the employees are.

    <!-- -->

    -   Complaints: Feel about how well complaints are handled by the management.

    <!-- -->

    -   Privileges: Feel about how the management allows for special privileges.

    <!-- -->

    -   Learning: Feel about to opportunities to learn new skills and grow within the role.

    <!-- -->

    -   Raises: Perception about pay raises are based on meritocracy rather than seniority or favoritism

    <!-- -->

    -   Critical: Feel about how much the department focuses on mistakes.

    <!-- -->

    -   Advance: Perception about promotion or advancement within the organization.

    The goal is to perform factorial analysis on this dataset and compare results from different methods of FA with Varimax rotation and also to interpret the most probable factors within the data.

    ```{r}
    library(psych) 
    data(attitude) 
    X <- attitude 
    n <- nrow(X) 
    p <- ncol(X) 
    Xs <- scale(X) 
    R <- 1/(n-1)*t(Xs)%*%Xs  
    res.eig <- eigen(R) 
    D <- diag(res.eig$values) 
    V <- res.eig$vectors
    ```

    1.  Compute the PCA approach to Factor Analysis:

    -   Decide the number of factors and compute the loadings.

    ```{r pcafa_factors_and_loadings}

    cumsum(res.eig$values)/sum(res.eig$values)

    plot(1:7, res.eig$values, type = "b")

    m <- 2

    L_PCA <- V[,1:m] %*% sqrt(D[1:m,1:m])
    L_PCA
    ```

    The first two components explain approximately 70% of the total variance, and the scree plot shows a clear elbow at the second eigenvalue. Therefore, we select **m = 2** factors.

    -   Compute communalities.

    ```{r pcafa_communalities}

    h <- apply(L_PCA, 1, function(row) sum(row^2))
    h
    ```

    -   Compute uniqueness (unique variances).

    ```{r pcafa_uniqueness}

    phi <- diag(diag(R-L_PCA %*% t(L_PCA)))
    phi
    ```

    -   Perform a varimax rotation (*?varimax*) and interpret the factors.

    ```{r pcafa_varimax}

    L_varimax <- varimax(L_PCA)

    L_varimax$loadings[1:p,1:m]
    ```

    Variables associated with **factor 1** are **rating, complaints, privilege** and **learning**. Variables associated with **factor 2** are **critical** and **advance**. Variable **raises** is associated with both factors.

    With this classification of variables it could be interpreted that **factor 1** captures aspects related to satisfaction of employees, perception of realization and perception of attention of employees needs from the management. In comparison, **factor 2** could be interpreted as the factor associated to perception of productivity and evaluation and recognition of work production. **Raises** could be interpreted as a variable that is related to satisfaction perception (**factor 1**) and work recognition (**factor 2**)

    -   Compute the factor scores by Thomson-Thurstone

    ```{r pcafa_scores}

    sigma <- L_PCA %*% t(L_PCA) + phi

    scores <- Xs %*% solve(sigma) %*% L_PCA
    scores
    ```

    2.  Compute the Maximum Likelihood Estimation approach to Factor Analysis using the `fa` method from the `psych` library using the argument **fm="ml"**:

    -   Decide the number of factors and provide the loadings.

    ```{r ml_factors_and_loadings}
    # ?fa

    paste("Deegres of freedom of discrepancy function n*F_ML")
    paste("m = ", 1:p, " df = ", 0.5*(p-1:p)^2-0.5*(p+1:p))

    res1 <- fa(Xs, nfactors = 1, fm = "ml")
    res2 <- fa(Xs, nfactors = 2, fm = "ml")
    res3 <- fa(Xs, nfactors = 3, fm = "ml")

    paste("p_value model m = 1: ", round(res1$PVAL,2))
    paste("p_value model m = 2: ", round(res2$PVAL,2))
    paste("p_value model m = 3: ", round(res3$PVAL,2))

    L <- res2$loadings[1:p,1:2]
    L
    ```

    As a starting point, it can be discarded m \>= 4 as provides negative values of degrees of freedom.

    From the possible choices the model m = 2 and 3 provides models with p-values greater than 0.05, meaning that the discrepancy between the model covariance matriz and the observed covariance matrix is not meaningful. And it will be picked m = 2 as is the simplest model between these two candidates.

    -   Give communalities.

    ```{r ml_communalities}
    res2$communalities
    ```

    -   Give the unique variances

    ```{r ml_unique_variances}
    res2$uniquenesses
    ```

    -   For each loading compute its sum of squares. Argument why communalities and squared sum of loadings are different.

    ```{r communalities_vs_ssl}

    h2 <- colSums(t(L)^2)
    h2
    ```

    The discrepancy is due to the assumption of correlation of factors. In the computation of sum of squares there is no assumption of correlation between factors, leading to inflated communalities. The results from res2\$communalities considers the correlation matrix as fa() allows correlated factors by default.

    -   Give the rotated loadings by a varimax rotation

    ```{r ml_varimax}

    res_varimax <- fa(Xs, nfactors = 2, fm = "ml", rotate = "varimax")
    res_varimax$loadings[1:p,1:2]
    ```

    -   Give the rotated loadings by a oblimin rotation

    ```{r ml_oblimin}

    res_oblimin <- fa(Xs, nfactors = 2, fm = "ml", rotate = "oblimin")
    res_oblimin$loadings[1:p,1:2]
    ```

    -   Compare the varimax rotated loadings with those of the oblimin. Interpret the factors

    ```{r ml_factor_interpretation}
    ```

    The rotated loadings obtained using **Varimax** (orthogonal rotation) and **Oblimin** (oblique rotation) are very similar. When contrasting results, the Oblimin rotation presents a slightly clearer and more separated structure for these variables compared to the Varimax rotation. However, the difference is not substantial. This can be interpreted to mean that while there may be a slight correlation between the underlying factors, it is not highly significant, as the Oblimin results do not differ drastically from the Varimax results (which assume zero correlation).

    In both cases, **Factor 1** loads strongly on **rating, complaints, privileges, learning, and raises**, representing a **general job satisfaction and management support**. **Factor 2** is clearly dominated by **advance**. This factor could be associated to perception of productivity and recognition of work production. The variable **raises** shows moderate loadings on both factors in both rotations.

    -   Compute the factor scores by Thomson

    ```{r ml_scores}
    L_varimax <- res_varimax$loadings
    L_oblimin <- res_oblimin$loadings
    sigma_varimax <- L_varimax %*% t(L_varimax) + diag(res_varimax$uniquenesses)

    sigma_oblimin <- L_oblimin %*% t(L_oblimin) + diag(res_oblimin$uniquenesses)

    scores_varimax <- Xs %*% solve(sigma_varimax) %*% L_varimax
    scores_oblimin <- Xs %*% solve(sigma_oblimin) %*% L_oblimin

    scores_varimax
    scores_oblimin
    ```

    -   Compare the rotated loadings from PCA and the ones from ML. Identify and discuss any significant discrepancies between the two sets of loadings.

    ```{r pcafa_vs_mlfa}
    df_comparison <- cbind(varimax(L_PCA)$loadings[1:p,1:2], 
            L_varimax[1:p,1:2],
            L_oblimin[1:p,1:2])

    colnames(df_comparison) <- c("f1_PCA", "f2_PCA", "f1_MLV", "f2_MLV", "f1_MLO", "f2_MLO")

    df_comparison
    ```

    Before any analysis it should be mentioned that the importance of comparison of loadings are in the magnitude not in the sign.

    The most important discrepancy between the two approaches is the difference in the loadings obtained from the PCA and ML methods. This difference can mainly be explained by the fact that PCA does not differentiate between shared variance and unique variance, whereas ML does.

    If, for instance, we focus on the *critical* variable, while in PCA it shows a strong loading on factor 2 (-0.73), under ML it drops to much weaker loadings (0.15 and 0.28). Observing the ML uniqueness for ***critical***, it has a value of 0.89, indicating that its shared variance is much lower than what PCA assumed, leading to weaker factor loadings. A similar reasoning can be applied to the rest of the variables, but *critical* presents the largest discrepancy between the two approaches.

    A second discrepancy point is observed in the variables ***privileges*****, *learning*,** and ***raises***. Under ML Varimax, these variables show cross-loadings (similar weights on both factors), making the interpretation and separation of the variables into latent factors more difficult. PCA shows a more distinct structure that facilitates this separation, though this might be an unrealistic representation due to the nature of the PCA method itself.
